Deep learning for quality control of surface physiographic fields using satellite Earth observations

### 2023-06-01 docs/submission/template.tex
Version of files used for resubmission

%\introduction  %% \introduction[modified heading if necessary]
\section{Introduction}
Accurate knowledge of the global surface physiography, including land, water and ice covers, and their characteristics, strongly determines the quality of surface and near-surface temperature simulations in weather and climate modelling. For instance, water bodies exchange mass and energy with the atmosphere and their thermal inertia strongly influence the lower boundary conditions such as skin temperatures, and surface fluxes of heat and moisture near the surface. Globally, there are $\sim$ 117 million lakes - defined as inland water bodies without lateral movement of water - making up around 3.7$\%$ of the Earth's land surface \citep{Verpoorter2014}. Their distribution is highly non-uniform, with the majority of lakes located between $45-75^{\circ}$N in the Boreal and Arctic regions. Lakes are highly important from the perspective of both numerical weather prediction and climate modelling as part of the EC-Earth model. For the latter, lakes generally influence the global carbon cycle as both sinks and sources of greenhouse gases; the majority of lakes are net heterotrophic (i.e. over saturated with carbon dioxide, CO$_2$), as a result of in lake respiration and so emit carbon into the atmosphere \citep{Pace2005,Tranvik2009}.  Total CO$_2$ emission from lakes is estimated at $1.25 - 2.30$ Pg of CO2-equivalents annually \citep{DelSontro2018}, nearly $20 \%$ of global CO$_2$ fossil fuel emissions, whilst lakes account for 9-24 $\%$  of CH$_4$ emissions, the second largest natural source after wetlands \citep{Saunois2020}. These rates of greenhouse gas emission are expected to rise further if the eutrophication (i.e. nutrient concentration increase) of the Earth's lentic systems continues. With regards to weather, freezing and melting of the lake surface modifies the radiative and conductive properties and consequently affects the heat (latent, sensible) exchange and surface energy balance \citep{Huang2019,Peng2020,Franz2018}. Considering particular examples, over Lake Victoria convective activity is suppressed during the day and peaks at night, leading to intense, hazardous thunderstorms \citep{Thiery2015,Thiery_2017}; Lake Ladoga can generate low level clouds which can cause variability in the 2m temperature of up to 10 K \citep{Eerola2014}; the Laurentian Great Lakes can cause intense winter snow storms \citep{Notaro2013,Vavrus2013}. Moreover, as a result of the increased temperatures due to climate change, lakes become more numerous due to the melting of glaciers and permafrost. Additionally, the higher temperatures mean that previously permanent lake bodies become seasonal or intermittent. There is then evidently a huge potential return in the ability to accurately model the location, morphology and properties of lakes in weather and climate models. \newline

@@

\noindent Lakes are incorporated in Earth-system models via parametrisation. At ECMWF the  representation of lakes via parametrisation was first handled by introducing the Fresh water Lake model FLake \citep{Mironov2008} into the IFS. FLake treats all resolved inland waterbodies (i.e. lakes, reservoirs, rivers which are dominating in a grid-cell) and unresolved or sub-grid water (i.e. small inland waterbodies and sea/ocean coastal waters which are present but not dominating in a grid-cell). Note that lake parameters are also an important part of the FLake model so when we refer in this work to ``lake parametrisation" we mean both the model and the parameters. The broad impact of the FLake model and the important role that waterbodies play in human life can be illustrated by analysing ECMWF maps of the fractional land sea mask and the inland waterbody cover alongside maps of the population density (i.e. inhabitants per km$^2$) based on the population count for 2015 from the Global Human Settlement Layers (GHSL), Population Grid 1975-2030 \citep{GHS,JRC100523} at 9 km horizontal resolution.

\hl{Globally FLake is active over 11.1\% of the grid-cells; considering only land grid-cells, then FLake is active over 32.4\% of the points. According to the population data, 64.5\% of densely populated areas (at least 300 inhabitants per km$^2$) are situated within a 9 km radius of a permanent waterbody (i.e. inland water or sea/ocean coast), with 31.2\% being in the vicinity of at least 1 km$^2$ waterbody - emphasising how essential waterbodies are in human life. In some regions this role may be even more crucial than in the others. For example in North America 45.7\% of the densely populated areas are close to a 1 km$^2$ waterbody; in Australia where only 0.5 \% of the land is populated,  two thirds of the population live within 9 km radius of a permanent waterbody of at least 1 km$^2$, with the majority of people living on the ocean coast.} \newline

%\hl{Globally FLake is active over 11.1\% of the grid-cells, with only 1.2\% of them being resolved inland waters (i.e. water covers $\geq$50\% of the grid-cell); considering only non-ocean (i.e. land) grid-cells, then FLake is active over 32.4\% of the grid-cells with only 3.5\% of them being resolved waters. According to the population data, only 4\% of land is densely populated (i.e. at least 300 inhabitants per km$^2$); 64.5\% of these areas being situated within a 9 km radius of a permanent waterbody (i.e. inland water or sea/ocean coast) with half of it (i.e. 31.2\% of densely populated areas) being in the vicinity of at least 1 km$^2$ waterbody - emphasising how essential waterbodies are in human life. In some regions this role may be even more crucial than in the others. For example, only 2\% of the North American region (similar for South American and North Asian regions) is densely populated with 45.7\% (33.9\% and 37.9\% respectively) of the areas being in vicinity of at least 1 km$^2$ waterbody; for Europe even though it has more densely populated areas (16\% of land is densely populated) still 37.4\% of the population are in the vicinity of at least a 1 km$^2$ waterbody; for a rather dry continent like Africa only 5\% of land is densely populated with 22.2\% of these areas being close to at least a 1 km$^2$ waterbody; most striking in this sense is Australia where only 0.5 \% of the land is populated, with two thirds of the population living within 9 km radius of a permanent waterbody of at least 1 km$^2$, with the majority of people living on the ocean coast.} \newline

\noindent It is a continuous enterprise to update the lake parametrization schemes and their input data fields to better represent small-scale surface processes. It is however challenging to accurately represent lakes in these parametrisations; the majority of lakes which are resolved at a 9km grid spacing have not had their morphology accurately measured, let alone monitored, whilst 28.9$\%$ of land and coastal cells are treated for sub-grid (i.e. covering half or less of a grid cell) water. When introducing an updated lake representation it is difficult apriori to determine the additional value gained through doing so. There are two key factors here:
\begin{itemize}
	\item Are the updated fields closer to reality?
	\item Do the updated fields increase the accuracy of the model predictions?
\end{itemize}
The first point is straightforward; we want our parametrisation fields to better represent reality. If the lake depth of some lake is updated from 10m to 100m we want to be sure that 100m is closer to the true depth of the lake. For the second point, even if the updated fields are accurate, are they informative in the sense that they enable us to make more accurate predictions? For instance, the main target of lake parametrization is to reproduce lake surface water temperatures (and therefore evaporation rates). If a lake parametrisation scheme is updated to better represent different types of inland waterbodies, the time variability of inland waterbodies and/or the lake morphology fields use more in situ measurements, does this additional information allow for more accurate predictions of the lake surface water temperatures? Is it therefore worthwhile to update the parametrisation in this way? Since the resulting updated fields are ultimately used operationally, it is essential to ensure the accuracy of the fields and prevent any potential degradation or instability of the model. This problem of quickly and automatically verifying the accuracy and information gain of updated lake parametrisations is the aim of this work. \newline

\noindent Numerical weather prediction and climate modelling are fields that are inherently linked with large datasets and complex, non-linear interactions. It is therefore an area that is particularly well placed to benefit from the deployment of machine learning algorithms. At ECMWF, advanced machine learning techniques have been used for parametrisation emulation via neural networks \citep{Chantry2021}, 4D-Var data assimilation \citep{Hatfield2021} and the post-processing of ensemble predictions \citep{Hewson2021}. Indeed, the early successes of these machine learning methods have led to the development of a 10-year roadmap for machine learning at ECMWF \citep{p19877}, with machine learning methods looking to be integrated into the operational workflow and machine learning demands considered in the procurement of HPC facilities. The ongoing development of novel computer architectures (e.g. GPU, IPU, FGPA) motivates utilizing algorithms and techniques which can efficiently take advantage of these new chips and gain significant performance returns. In this work we will demonstrate a new technique for the Verification of Earth-System ParametERisation (VESPER) based on a deep learning neural network regression model. This tool enables the accuracy of an updated water body parametrization to be rapidly and automatically assessed, and the added value that such an updated parametrization brings to be quantitatively evaluated. \newline

\noindent This paper is organized as follows. In Section \ref{sec:2} we describe the construction of the VESPER tool - the raw input data, the processing steps and the construction of a neural network regressor. In Section \ref{sec:3} we then deploy VESPER to investigate and evaluate updated lake parametrisation fields. Discussion and concluding remarks are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively.

\section{Constructing VESPER}\label{sec:2}
In order to rapidly assess the accuracy of new surface physiography fields and if their use in the model increase the accuracy with which we can make predictions, a neural network regression model (VESPER, hereafter) that can learn the mapping between a set of input features $x$ and targets $y$ is constructed. In this case the features are the atmospheric and surface model fields (such as 2 metre temperature from ERA5 reanalysis) and the surface physiographic fields (such as orography and vegetation cover used to produce ERA5 reanalysis). See Table \ref{table:definitions} for the full list of variables used. The target is the satellite land surface temperature (skin temperature from MODIS Aqua Day MYD11A1 v006 collection). Once trained, VESPER can then make predictions about the skin temperature given a set of input variables (i.e. atmospheric and surface model fields, and surface physiographic fields). In turn, these predictions can then be compared against observations (i.e. satellite skin temperature) and VESPER's accuracy evaluated. By varying the number, type and values of the input features to VESPER and observing how the accuracy of its predictions change, some conclusions on if and how features can increase predictability of an actual atmospheric model can be drawn. Moreover, by isolating geographic regions where the predictions get worse with new/updated surface physiographic fields, areas where these fields might be erroneous or not informative enough can be identified. Due to the inherent stochasticity of training a neural network regression model it is also possible for different models to settle in different local minimums i.e. the network variance/noise. To understand the significance of this, every VESPER configuration was trained four times, each time with a different random seed. \newline

@@

\subsection{Features and targets}
VESPER's input feature selection (see Table \ref{table:definitions}) followed (i) permutation importance results for atmospheric and surface model fields - only fields with the highest importance were chosen; and (ii) expert choice for surface physiographic fields. As a first attempt it was decided to test the current methodology for lake related information, therefore fields that could be most affected by the presence of absence of water were selected, e.g. if lake had to be removed then some other surface had to appear (like bare ground, high or low vegetation, glacier or even ocean) and surface elevation had to change. Changes to the orographic fields will have important influences on temperature through e.g. wind, solar heating etc. Lake depth changes are similarly important, influencing how a lake freezes, thaws, mixes and its overall dynamical range. VESPER's target selection followed globally available criteria and the satellite land surface temperature is quite well observed globally and with high temporal pattern (daily or even several times a day depending on the location)

@@

There are three main sources of data. The first is selection of surface physiographic fields from ERA5 \citep{Hersbach} and their updated versions \citep{Choulga2019,Boussetta2021,Mu2021} used as VESPER’s features. As a shorthand we will refer to the original ERA5 physiographic fields as version ``V15" and the updated versions as ``V20". The second is a selection of atmospheric and surface model fields from ERA5, also used as VESPER’s features. The third is day-time land surface temperature measurements from the Moderate Resolution Imaging Spectroradiometer (MODIS) onboard the Aqua satellite \citep{MODIS}, used as VESPER’s target variable.

\subsubsection{Surface physiographic fields}\label{sec:surface_physio}
\noindent Surface physiographic fields have gridded information of the Earth’s surface properties (e.g. land-use, vegetation type and distribution) and represent surface heterogeneity in the ECLand of ECMWF’s Integrated Forecasting System (IFS). They are used to compute surface turbulent fluxes (of heat, moisture and momentum) and skin temperature over different surfaces (vegetation, bare soil, snow, interception and water) and then to calculate an area-weighted average for the grid-box to couple with the atmosphere. To trigger all different parametrization schemes the ECMWF model uses a sets of physiographic fields, that do not depend on initial condition or forecast step. Most fields are constant; surface albedo is specified for 12 months to describe the seasonal cycle. Dependent on the origin, initial data comes at different resolutions and different projections, and is then first converted to a regular latitude-longitude grid (EPSG:4326) at $\sim$ 1km at equator resolution, and secondly to a required grid and resolution. Surface physiographic fields used in this work consist of orographic, land, water, vegetation, soil, albedo fields and their difference between initial V15 and updated V20 field sets. See Tables \ref{table:definitions} and \ref{tab:datasources} for the full list of surface physiographic fields and their input sources; for more details see IFS documentation \citep{IFSdocs}. As this work is focused on assessing quality of inland water information, main surface physiographic fields are lake cover (derived from land-sea mask) and lake mean depth (see Table \ref{tab:datasources}).  \newline 
\begin{table*}
	\begin{tabularx}{\textwidth}{lXX}
		\toprule 
		Field Category & V15 & V20 \\
		Orographic & SRTM30 Shuttle Radar Topography Mission over 60°N-60°S; GLOBE: Global Land One-km Base Elevation Project data over 90-60°N; RAMP2: high-resolution Radarsat Antarctic Mapping Project Digital Elevation Model version 2 data \citep{Liu2015} over 60-90°S; BPRC: Byrd Polar Research Center over Greenland; IS 50V: Digital Map Database of Iceland over Iceland & As V15, with corrections to to erroneous shift
		\\
		Land   & \textbf{Source lsm:} GlobCover2009 \citep{GLOBCOVER,arino2012glcm} over 85°N-60°S; RAMP2: high-resolution Radarsat Antarctic Mapping Project Digital Elevation Model version 2 data \citep{Liu2015} over 60-90°S; no land assumed over 90-85°N \newline  
		\textbf{Source glm:} GLCC: Global Land Cover Characteristics version 2.0 over 90°N-90°S except Iceland; Icelandic Meteorological Office (IMO) glacier mask 2013 over Iceland
		 & \textbf{Source lsm:} GSWE: Global Surface Water Explorer \citep{GSWE}; glm \newline 
		 \textbf{Source glm:} Norwegian Institute glacier data over Svalbard; Icelandic Meteorological Office (IMO) glacier mask 2017 over Iceland; GIMP: Greenland Ice Mapping Project data \citep{Howat2014} over Greenland; CryoSat-2 satellite glacier data \citep{Slater2018} over Antarctica (+ manual gap filling);  GLIMS: Global Land Ice Measurements from Space data \citep{glims} over rest of the globe \\

@@

				\textbf{Source cl:} \textit{lsm} (ocean is separated at actual resolution by seeding and removing all connected grid-cells, includes the Caspian Sea, the Azov Sea, The American Great Lakes) \newline 
				\textbf{Source dl:} The Caspian Sea bathymetry; Global Relief Model ETOPO1 \citep{amante2009egrm} over the Great Lakes, the Azov Sea; GLDB: Global Lake DataBase version 1 \citep{Kourzeneva2012} over rest of the globe; 25 meters assumed over missing data grid-cells
		& \textbf{Source cl:} \textit{lsm} (ocean is separated at ~1km resolution by upgraded flooding algorithm following \cite{Choulga2019} \newline 
		\textbf{Source dl:} GEBCO: General Bathymetric Charts of the Ocean \citep{Weatherall2015} over the Caspian Sea and the Azov Sea; Global Relief Model ETOPO1 \citep{amante2009egrm} over the Great Lakes; GLDB: Global Lake DataBase version 3 \citep{Choulga2014} over rest of the globe; indirect estimates based on geological origin of lakes \citep{Choulga2014} over missing data grid-cells \\

@@

		 Albedo & MODIS 5-year climatology \citep{SCHAAF2002135}; RossThickLiSparseReciprocal BRDF model. note that Albedo values represent snow free surface albedo & As V15 \\ 

@@

	\caption{List of input datasets for the surface physiographic fields for V15 and V20 field sets. V15X and V20X are identical to V15 and V20 respectively, but with the addition of saline lake cover, and monthly varying lake cover fields.}

@@

To expand V15 and V20 lake description (to V15X and V20X respectively, see Table \ref{tab:datasources}) their salinity and time variability information was generated. Even though static permanent water fits better to describe inland water distribution on average all year round, some areas (in Tropics especially) could  benefit from having monthly varying information as they have a very strong seasonal cycle, when size, shape and depth of a lake changes over the course of the year, leading to a significant change in modelling the lake temperature response. Similarly, saline lakes behave very differently to fresh water lakes since increased salt concentrations affect the density, specific heat capacity, thermal conductivity, and turbidity, as well as evaporation rates, ice formation and ultimately the surface temperature. These two properties of time variability and salinity are often related; it is common for saline lakes to fill and dry out over the course of the season, which naturally also affects the relative saline concentration of the lake itself. To create a monthly varying lake cover first 12 monthly fractional land-sea masks based on JRC Monthly Water History v1.3 maps for 2010-2020 were created. Since the annual lake maps were created taking into account a lot of additional sources the extra condition on the monthly maps that the monthly water is equal or greater than permanent water distribution from fractional land-sea mask is enforced. To create an inland salt lake cover map, the GLDBv3 salt lake list was used. First, in order to identify separate lakes on $\sim$ 1km resolution lake cover, small sub-grid lakes and large lake coasts are masked, i.e. grid-cells that have water fraction less than 0.25. Next, number of connected grid-cells in each lake (i.e. connected with sides only) is computed. Then only lakes that have 100 and more connected grid-cells are vectorised, as at ERA5 resolution of ∼31km the grid-cells are quite large and can include a mixture of freshwater and saline lakes. Finally, saline lake vectors are selected by filtering vectors which have no saline lake point from GLDBv3 located – in total 147 large salt lake vectors, which were further used to filter non-saline lakes at ~1km resolution lake cover, finally aggregated to ~31km resolution. In the future it is planned to revisit this field and extend the list to include additional data. Note that all non-lake related climate fields such as vegetation cover or orography were updated in V20 field set compared to V15 only in relation to the changing lake fields (i.e. if fraction of lake in the grid cell increased then other fractions like vegetation or bare ground should have increased accordingly).

\begin{figure}
	\subfloat[\label{fig:cl_map}]{\includegraphics[width=0.48\textwidth]{cl_map}} \\
	\subfloat[\label{fig:cl_delta_map}]{\includegraphics[width=0.48\textwidth]{cl_delta_map}}
	\caption{At ~31km resolution (a) V20 fractional lake cover and (b) difference between V20 and V15 lake covers. Over northern latitudes inland water increase in V20 compared to V15 is due to higher resolution input source and its better satellite image recognition methodologies; inland water reduction in V20 compared to V15 is due to anthropogenic land use changes (e.g. Aral Sea) or due to use of only permanent water (e.g. Australia) which was proven to better represent inland water distribution on annual basis.} 
	\label{fig:example_figure_a}
\end{figure}

>>>>

%\introduction  %% \introduction[modified heading if necessary]
\section{Introduction}
Accurate knowledge of the global surface physiography, including land, water and ice covers, and their characteristics, strongly determines the quality of surface and near-surface temperature simulations in weather and climate modelling. For instance, water bodies exchange mass and energy with the atmosphere and their thermal inertia strongly influence the lower boundary conditions such as skin temperatures, and surface fluxes of heat and moisture near the surface. Globally, there are $\sim$ 117 million lakes - defined as inland water bodies without lateral movement of water - making up around 3.7$\%$ of the Earth's land surface \citep{Verpoorter2014}. Their distribution is highly non-uniform, with the majority of lakes located between $45-75^{\circ}$N in the Boreal and Arctic regions. Lakes are highly important from the perspective of both numerical weather prediction and climate modelling as part of the EC-Earth model. For the latter, lakes generally influence the global carbon cycle as both sinks and sources of greenhouse gases; the majority of lakes are net heterotrophic (i.e. over saturated with carbon dioxide, CO$_2$), as a result of in lake respiration and so emit carbon into the atmosphere \citep{Pace2005,Tranvik2009}.  Total CO$_2$ emission from lakes is estimated at $1.25 - 2.30$ Pg of CO$_2$-equivalents annually \citep{DelSontro2018}, nearly $20 \%$ of global CO$_2$ fossil fuel emissions, whilst lakes account for 9-24 $\%$  of CH$_4$ emissions, the second largest natural source after wetlands \citep{Saunois2020}. These rates of greenhouse gas emission are expected to rise further if the eutrophication (i.e. nutrient concentration increase) of the Earth's lentic systems continues. With regards to weather, freezing and melting of the lake surface modifies the radiative and conductive properties and consequently affects the heat (latent, sensible) exchange and surface energy balance \citep{Franz2018,Huang2019,Peng2020}. Considering particular examples, over Lake Victoria convective activity is suppressed during the day and peaks at night, leading to intense, hazardous thunderstorms \citep{Thiery2015,Thiery_2017}; Lake Ladoga can generate low level clouds which can cause variability in the 2m temperature of up to 10 K \citep{Eerola2014}; the Laurentian Great Lakes can cause intense winter snow storms \citep{Notaro2013,Vavrus2013}. Moreover, as a result of the increased temperatures due to climate change, lakes become more numerous due to the melting of glaciers and permafrost. Additionally, the higher temperatures mean that previously permanent lake bodies become seasonal or intermittent. There is then evidently a huge potential return in the ability to accurately model the location, morphology and properties of lakes in weather and climate models. \newline

@@

\noindent Lakes are incorporated in Earth-system models via parametrisation. At ECMWF the  representation of lakes via parametrisation was first handled by introducing the Fresh water Lake model FLake \citep{Mironov2008} into the IFS. FLake treats all resolved inland waterbodies (i.e. lakes, reservoirs, rivers which are dominating in a grid-cell) and unresolved or sub-grid water (i.e. small inland waterbodies and sea/ocean coastal waters which are present but not dominating in a grid-cell). Its main drivers (input fields) are lake location and lake mean depth. The broad impact of the FLake model (i.e. areas where it is active) and the important role that waterbodies play in human life can be illustrated by analysing ECMWF fields of the fractional land sea mask and the inland waterbody cover alongside the population density field (i.e. inhabitants per km$^2$) based on the population count for 2015 from the Global Human Settlement Layers (GHSL), Population Grid 1975-2030 \citep{JRC100523,GHS} at 9 km horizontal resolution.

%
%Note that lake parameters are also an important part of the FLake model so when we refer in this work to ``lake parametrisation" we mean both the model and the parameters.

Globally FLake is active over 11.1\% of the grid-cells; considering only land grid-cells, then FLake is active over 32.4\% of the points. According to the population data, 64.5\% of densely populated areas (at least 300 inhabitants per km$^2$) are situated within a 9 km radius of a permanent waterbody (i.e. inland water or sea/ocean coast), with 31.2\% being in the vicinity of at least 1 km$^2$ waterbody - emphasising how essential waterbodies are in human life. In some regions this role may be even more crucial than in the others. For example in North America 45.7\% of the densely populated areas are close to a 1 km$^2$ waterbody; in Australia where only 0.5 \% of the land is populated,  two thirds of the population live within 9 km radius of a permanent waterbody of at least 1 km$^2$, with the majority of people living on the ocean coast. \newline

\noindent It is a continuous enterprise to update the lake parametrization input fields to better represent small-scale surface processes. It is however challenging to do it accurately as the majority of lakes which are resolved at a 9km grid spacing have not had their morphology accurately measured, let alone monitored, whilst 28.9$\%$ of land and coastal cells are treated for sub-grid (i.e. covering half or less of a grid cell) water. When introducing an updated lake representation it is difficult apriori to determine the additional value gained through doing so. There are two key factors here:
\begin{itemize}
	\item Are the updated fields closer to reality?
	\item Do the updated fields increase the accuracy of the model predictions?
\end{itemize}
The first point is straightforward; we want our fields to better represent reality. If the lake depth of some lake is updated from 10m to 100m we want to be sure that 100m is closer to the true depth of the lake. For the second point, even if the updated fields are accurate, are they informative in the sense that they enable us to make more accurate predictions? For instance, the main target of lake parametrization is to reproduce lake surface water temperatures (and therefore evaporation rates). If lake parametrisation input fields are updated to better represent different types of inland waterbodies, the time variability of inland waterbodies and/or the lake morphology fields use more in situ measurements, does this additional information allow for more accurate predictions of the lake surface water temperatures? Is it therefore worthwhile to spend several person-months to update/create a lake-related field? Since the resulting updated fields are ultimately used operationally, it is essential to ensure the accuracy of the fields and prevent any potential degradation or instability of the model. This problem of quickly and automatically checking the accuracy and information gain of updated lake-related fields is the aim of this work. \newline

\noindent Numerical weather prediction and climate modelling are domains that are inherently linked with large datasets and complex, non-linear interactions. It is therefore an area that is particularly well placed to benefit from the deployment of machine learning algorithms. At ECMWF, advanced machine learning techniques have been used for parametrisation emulation via neural networks \citep{Chantry2021}, 4D-Var data assimilation \citep{Hatfield2021} and the post-processing of ensemble predictions \citep{Hewson2021}. Indeed, the early successes of these machine learning methods have led to the development of a 10-year roadmap for machine learning at ECMWF \citep{p19877}, with machine learning methods looking to be integrated into the operational workflow and machine learning demands considered in the procurement of HPC facilities. The ongoing development of novel computer architectures (e.g. GPU, IPU, FGPA) motivates utilizing algorithms and techniques which can efficiently take advantage of these new chips and gain significant performance returns. In this work we will demonstrate a new technique for the Verification of Earth-System ParametERisation (VESPER) based on a deep learning neural network regression model. This tool enables the accuracy of an updated water body-related field to be rapidly and automatically assessed, and the added value that such updated fields bring to be quantitatively evaluated. \newline

\noindent This paper is organized as follows. In Section \ref{sec:2} we describe the construction of the VESPER tool - the raw input data, the processing steps and the construction of a neural network regressor. In Section \ref{sec:3} we then deploy VESPER to investigate and evaluate updated lake-related fields. Discussion and concluding remarks are made in Sections \ref{sec:discussion} and \ref{sec:conclusion} respectively.

\section{Constructing VESPER}\label{sec:2}
In order to rapidly assess the accuracy of new surface physiography fields and if their use in the model increase the accuracy with which we can make predictions, a neural network regression model (VESPER, hereafter) that can learn the mapping between a set of input features $x$ and targets $y$ is constructed. In this case the features are the atmospheric and surface model fields (such as 2 metre temperature from ERA5 reanalysis) and the surface physiographic fields (such as orography and vegetation cover used to produce ERA5 reanalysis). See Table \ref{table:definitions} for the full list of variables used. The target is the satellite land surface temperature (LST; skin temperature from MODIS Aqua Day MYD11A1 v006 collection). Once trained, VESPER can then make predictions about the skin temperature given a set of input variables (i.e. atmospheric and surface model fields, and surface physiographic fields). In turn, these predictions can then be compared against observations (i.e. satellite skin temperature) and VESPER's accuracy evaluated. By varying the number, type and values of the input features to VESPER and observing how the accuracy of its predictions change, some conclusions on if and how features can increase predictability of an actual atmospheric model can be drawn. Moreover, by isolating geographic regions where the predictions get worse with new/updated surface physiographic fields, areas where these fields might be erroneous or not informative enough can be identified. Due to the inherent stochasticity of training a neural network regression model it is also possible for different models to settle in different local minimums i.e. the network variance/noise. To understand the significance of this, every VESPER configuration was trained four times, each time with a different random seed. \newline

@@

\subsection{Features and targets}
VESPER's input feature selection (see Table \ref{table:definitions}) followed (i) permutation importance results for atmospheric and surface model fields - only fields with the highest importance were chosen; and (ii) expert choice for surface physiographic fields. As a first attempt it was decided to test the current methodology for lake related information, therefore fields that could be most affected by the presence or absence of water were selected, e.g. if lake had to be removed then some other surface had to appear (like bare ground, high or low vegetation, glacier or even ocean) and surface elevation had to change. Changes to the orographic fields will have important influences on temperature through e.g. wind, solar heating, etc. Lake depth changes are similarly important, influencing how a lake freezes, thaws, mixes and its overall dynamical range. VESPER's target selection followed globally available criteria and the satellite LST is quite well observed globally and with high temporal pattern (daily or even several times a day depending on the location).

@@

There are three main sources of data. The first is selection of surface physiographic fields from ERA5 \citep{Hersbach} and their updated versions \citep{Choulga2019,Boussetta2021,Mu2021} used as VESPER’s features. As a shorthand we will refer to the original ERA5 physiographic fields as version ``V15" and the updated versions as ``V20". The second is a selection of atmospheric and surface model fields from ERA5, also used as VESPER’s features. The third is day-time LST measurements from the Moderate Resolution Imaging Spectroradiometer (MODIS) onboard the Aqua satellite \citep{MODIS}, used as VESPER’s target variable.

\subsubsection{Surface physiographic fields}\label{sec:surface_physio}
\noindent Surface physiographic fields have gridded information of the Earth’s surface properties (e.g. land-use, vegetation type and distribution) and represent surface heterogeneity in the ECLand of the IFS. They are used to compute surface turbulent fluxes (of heat, moisture and momentum) and skin temperature over different surfaces (vegetation, bare soil, snow, interception and water) and then to calculate an area-weighted average for the grid-box to couple with the atmosphere. To trigger all different parametrization schemes the ECMWF model uses a sets of physiographic fields, that do not depend on initial condition or forecast step. Most fields are constant; surface albedo is specified for 12 months to describe the seasonal cycle. Dependent on the origin, initial data comes at different resolutions and different projections, and is then first converted to a regular latitude-longitude grid (EPSG:4326) at $\sim$ 1km at Equator resolution, and secondly to a required grid and resolution. Surface physiographic fields used in this work consist of orographic, land, water, vegetation, soil, albedo fields and their difference between initial V15 and updated V20 field sets. See Tables \ref{table:definitions} and \ref{tab:datasources} for the full list of surface physiographic fields and their input sources; for more details see IFS documentation \citep{IFSdocs}. As this work is focused on assessing quality of inland water information, main surface physiographic fields are lake cover (derived from land-sea mask) and lake mean depth (see Table \ref{tab:datasources}).  \newline 
\begin{table*}
	\begin{tabularx}{\textwidth}{lXX}
		\toprule 
		Field category & V15 (initial) & V20 (updated) \\
		\hline
		Orographic & SRTM30 Shuttle Radar Topography Mission over 60°N-60°S; GLOBE: Global Land One-km Base Elevation Project data over 90-60°N; RAMP2: high-resolution Radarsat Antarctic Mapping Project Digital Elevation Model version 2 data \citep{Liu2015} over 60-90°S; BPRC: Byrd Polar Research Center over Greenland; IS 50V: Digital Map Database of Iceland over Iceland & As V15, with corrections of erroneous shift
		\\
		\hline 
			Land &	\textbf{glm:} GLCC: Global Land Cover Characteristics version 2.0 over 90°N-90°S except Iceland; Icelandic Meteorological Office (IMO) glacier mask 2013 over Iceland \newline 
		\textbf{lsm:} GlobCover2009 \citep{GLOBCOVER,arino2012glcm} over 85°N-60°S; RAMP2: high-resolution Radarsat Antarctic Mapping Project Digital Elevation Model version 2 data \citep{Liu2015} over 60-90°S; no land assumed over 90-85°N

		 &  \textbf{glm:} Norwegian Institute glacier data over Svalbard; Icelandic Meteorological Office (IMO) glacier mask 2017 over Iceland; GIMP: Greenland Ice Mapping Project data \citep{Howat2014} over Greenland; CryoSat-2 satellite glacier data \citep{Slater2018} over Antarctica (+ manual gap filling);  GLIMS: Global Land Ice Measurements from Space data \citep{glims} over rest of the globe  \newline 
		 \textbf{lsm:} GSWE: Global Surface Water Explorer \citep{GSWE}; glm \\

@@

				\textbf{cl:} \textit{lsm} (ocean is separated at actual resolution by seeding and removing all connected grid-cells, includes the Caspian Sea, the Azov Sea, The American Great Lakes) \newline 
				\textbf{dl:} The Caspian Sea bathymetry; Global Relief Model ETOPO1 \citep{amante2009egrm} over the Great Lakes, the Azov Sea; GLDB: Global Lake DataBase version 1 \citep{Kourzeneva2012} over rest of the globe; 25 meters assumed over missing data grid-cells
		& \textbf{cl:} \textit{lsm} (ocean is separated at ~1km resolution by upgraded flooding algorithm following \cite{Choulga2019} \newline 
		\textbf{dl:} GEBCO: General Bathymetric Charts of the Ocean \citep{Weatherall2015} over the Caspian Sea and the Azov Sea; Global Relief Model ETOPO1 \citep{amante2009egrm} over the Great Lakes; GLDB: Global Lake DataBase version 3 \citep{Choulga2014} over rest of the globe; indirect estimates based on geological origin of lakes \citep{Choulga2014} over missing data grid-cells \\

@@

		 Albedo & MODIS 5-year climatology \citep{SCHAAF2002135}; RossThickLiSparseReciprocal BRDF model. Note that Albedo values represent snow free surface albedo & As V15 \\ 

@@

	\caption{List of input datasets for the surface physiographic fields for V15 and V20 field sets. V15X and V20X are identical to V15 and V20 respectively, but with the addition of saline lake cover, and monthly varying lake cover fields. \newline }

@@

To expand V15 and V20 lake description (to V15X and V20X respectively) their salinity and time variability information was generated. Even though static permanent water fits better to describe inland water distribution on average all year round, some areas (in Tropics especially) could  benefit from having monthly varying information as they have a very strong seasonal cycle, when size, shape and depth of a lake changes over the course of the year, leading to a significant change in modelling the lake temperature response. Similarly, saline lakes behave very differently to fresh water lakes since increased salt concentrations affect the density, specific heat capacity, thermal conductivity, and turbidity, as well as evaporation rates, ice formation and ultimately the surface temperature. These two properties of time variability and salinity are often related; it is common for saline lakes to fill and dry out over the course of the season, which naturally also affects the relative saline concentration of the lake itself. To create a monthly varying lake cover first 12 monthly fractional land-sea masks based on JRC Monthly Water History v1.3 maps for 2010-2020 were created. Since the annual lake maps were created taking into account a lot of additional sources the extra condition on the monthly maps that the monthly water is equal or greater than permanent water distribution from fractional land-sea mask is enforced. To create an inland salt lake cover map, the GLDBv3 salt lake list was used. First, in order to identify separate lakes on $\sim$ 1km resolution lake cover (by ``lake cover" we refer the maximum lake distribution based on 12 monthly-varying lake covers), small sub-grid lakes and large lake coasts are masked, i.e. grid-cells that have water fraction less than 0.25. Next, number of connected grid-cells in each lake (i.e. connected with sides only) is computed. Then only lakes that have 100 and more connected grid-cells are vectorised, as at ERA5 resolution of ∼31km the grid-cells are quite large and can include a mixture of freshwater and saline lakes. Finally, saline lake vectors are selected by filtering vectors which have no saline lake point from GLDBv3 located – in total 147 large salt lake vectors, which were further used to filter non-saline lakes at ~1km resolution lake cover, finally aggregated to ~31km resolution. In the future it is planned to revisit this field and extend the list to include additional data. Note that all non-lake related climate fields such as vegetation cover or orography were updated in V20 field set compared to V15 only in relation to the changing lake fields (i.e. if fraction of lake in the grid cell increased then other fractions like vegetation or bare ground should have increased accordingly).

\begin{figure}
	\subfloat[\label{fig:cl_map}]{\includegraphics[width=0.48\textwidth]{cl_map}} \\
	\subfloat[\label{fig:cl_delta_map}]{\includegraphics[width=0.48\textwidth]{cl_delta_map}}
	\caption{At $\sim$ 31km resolution (a) V20 fractional lake cover and (b) difference between V20 and V15 lake covers. Over northern latitudes inland water increase in V20 compared to V15 is due to higher resolution input source and its better satellite image recognition methodologies as well as thawing permafrost; inland water reduction in V20 compared to V15 is due to anthropogenic land use changes (e.g. Aral Sea) or due to use of only permanent water (e.g. Australia) which was proven to better represent inland water distribution on annual basis.} 
	\label{fig:example_figure_a}
\end{figure}
